How is experience transformed into memory?

### 2018-09-01 paper/main.tex
backing up edits to results

\section*{Results}
To characterize the gists of the \textit{Sherlock} episode participants watched and their subsequent recountings of the episode, we used a topic model~\cite{BleiEtal03} to discover the latent thematic content in the video.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents; they return as output two matrices.  The first output is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes which mix topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 45 minute video.  This information included: a brief narrative description of what was happening; whether the scene took place indoors vs. outdoors; names of any characters on the screen; names of any characters who were in focus in the camera show; names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annoted features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purposes of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 28 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the movie (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize the how the properties of collections of documents change over time, our approach allows us to examine the topic dynamics within a single episode.  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

The topics we found were heavily character-focused (e.g., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes related to primarily Sherlock-focused; primarily (John) Watson-focused (Watson is Sherlock's close confidant and assistant); and themes that involved Sherlock and Watson interacting (Fig.~\topics).  Several of the topics also appeared highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences (if the distinctions between those overlapping topics were meaningful).  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) topics tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistance} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B).  Following \cite{BaldEtal17}, used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow).  Part of our model fitting procedure required selecting an appropriate number of ``events'' to segment the timeseries into.  We used an optimization procedure to identify the number of events that maximized within-event stability while also minimizing across-event correlations (see \textit{Methods} for additional details).  To create a stable ``summary'' of the video, we computed the avereage topic vector within each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants recalls we treated each (overlapping) 10 sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a consequence of projecting the video and recalls into a common ``topic'' space.  An example matrix from one participant is shown in Figure~\ref{fig:model}D.

To

%JRM STOPPED HERE

% characterize gist of movie and recalls
% x1. movie: fit topic models (add note on dynamic topic models).  fig 1: what do the topics reflect (top words from each topic), example with a few scenes.
% x2. notice block diagonal structure, rather than smooth transitions; motivates HMM approach.  Note about selecting K
% 3. question: do recalls exhibit same structure?  apply topic model to recalls.  note that by projecting everything into a common topic space, the thematic content of each moment of movie/recall can be compared, even though the mix of words used to describe any given event or moment may not be consistent across people (or across annotations/recall)
% 4. recalls exhibit blocky structure, but at different scales.  motivates fitting K separately by person
% 5. match up events (correlations)
% 6. now we can use classic memory analyses to examine structure of recall.  motivate and reference PFR, lag-CRP, and SPC.  but this does not tell us about structure in the *content* of what people remembered, or whether people's recalls are capturing the hierarchical temporal structure of the original movie.
% 7. topic trajectories as a new way of characterizing recall.  assess agreement using rayleigh test.  also examine shapes of each subject's trajectory to show match.  key insight: everyone captures the general shape, but at different "sampling resolutions"
% 8. we can also project back onto the original (word) features to understand more about what any particular part of the trajectory reflects (wordle fig)
% 9. the above analyses tell us about how the trajectory of thematic content in the episode is both preserved and transformed in participants' memories.  we next carried out analyses to understand the neural bases of these phenomena.  analysis 1: which structures are sensitive (during viewing) to the original episode's topic trajectory?  answer: default mode.  anlaysis 2: which structure are sensitive (during viewing) to how the participants will later recall the episode?  answer: vmPFC, default mode.  (potential role of vmPFC in reflecting participant-specific memory transformations and processing)

>>>>

\section*{Results}
To characterize the gists of the \textit{Sherlock} episode participants watched and their subsequent recountings of the episode, we used a topic model~\citep{BleiEtal03} to discover the latent thematic content in the video.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents; they return as output two matrices.  The first output is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes which mix topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 45 minute video.  This information included: a brief narrative description of what was happening; whether the scene took place indoors vs. outdoors; names of any characters on the screen; names of any characters who were in focus in the camera; names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annoted features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purposes of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 28 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the movie (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize the how the properties of collections of documents change over time, our approach allows us to examine the topic dynamics within a single video.  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

The topics we found were heavily character-focused (e.g., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes that were primarily Sherlock-focused; primarily (John) Watson-focused (Watson is Sherlock's close confidant and assistant); or that involved Sherlock and Watson interacting (Fig.~\topics).  Several of the topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences (if the distinctions between those overlapping topics were meaningful).  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) topics tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistance} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B).  Following \cite{BaldEtal17}, we used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow).  Part of our model fitting procedure required selecting an appropriate number of ``events'' to segment the timeseries into.  We used an optimization procedure to identify the number of events that maximized within-event stability while also minimizing across-event correlations (see \textit{Methods} for additional details).  To create a stable ``summary'' of the video, we computed the avereage topic vector within each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls we treated each (overlapping) 10 sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a huge benefit of projecting the video and recalls into a common ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as for the video (e.g., compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (e.g., most topics are active or inactive over contiguous blocks of time), the overall timecourses are not as cleanly delineated as the video topics are.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example partipant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we can use an HMM, along with the aformentioned event number optimization procedure (also see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (e.g.\ every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g.\ Participants 1, 4, and 15), while others' recalls segmented into many shorter duration events (e.g.\ Participants 12, 13, and 17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g. some might touch on just the major plot points, whereas others might attempt to recall every minor scene.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations in participant's recalls.  Whereas each event in the original video (was largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory participants appear to be integrating \textit{across} different events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event (Figs.~\ref{fig:model}D, \corrmats).

\textbf{JRM STOPPED HERE}

% characterize gist of movie and recalls
% x1. movie: fit topic models (add note on dynamic topic models).  fig 1: what do the topics reflect (top words from each topic), example with a few scenes.
% x2. notice block diagonal structure, rather than smooth transitions; motivates HMM approach.  Note about selecting K
% x3. question: do recalls exhibit same structure?  apply topic model to recalls.  note that by projecting everything into a common topic space, the thematic content of each moment of movie/recall can be compared, even though the mix of words used to describe any given event or moment may not be consistent across people (or across annotations/recall)
% x4. recalls exhibit blocky structure, but at different scales.  motivates fitting K separately by person
% 5. match up events (correlations)
% 6. now we can use classic memory analyses to examine structure of recall.  motivate and reference PFR, lag-CRP, and SPC.  but this does not tell us about structure in the *content* of what people remembered, or whether people's recalls are capturing the hierarchical temporal structure of the original movie.
% 7. topic trajectories as a new way of characterizing recall.  assess agreement using rayleigh test.  also examine shapes of each subject's trajectory to show match.  key insight: everyone captures the general shape, but at different "sampling resolutions"
% 8. we can also project back onto the original (word) features to understand more about what any particular part of the trajectory reflects (wordle fig)
% 9. the above analyses tell us about how the trajectory of thematic content in the episode is both preserved and transformed in participants' memories.  we next carried out analyses to understand the neural bases of these phenomena.  analysis 1: which structures are sensitive (during viewing) to the original episode's topic trajectory?  answer: default mode.  anlaysis 2: which structure are sensitive (during viewing) to how the participants will later recall the episode?  answer: vmPFC, default mode.  (potential role of vmPFC in reflecting participant-specific memory transformations and processing)
